---
title: "Intro to Transcriptomics Part II: DESeq2"
author: Nicole Black, Kayla Xu
date: 07/30/22
updated: 02/26/23
---

```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("/Users/matthewgenchev/Desktop/USC/qbio_490_matthewgenchev/analysis_data")) 
```

```{r}
#install "DESeq2" using BiocManager::install()
#BiocManager::install("DESeq2")
# load DESEq2

library(DESeq2)
```

```{r}
# Read in csv files

rna_counts <- read.csv("/Users/matthewgenchev/Desktop/USC/qbio_490_matthewgenchev/analysis_data/brca_rna_counts_data.csv")
rna_genes <- read.csv("/Users/matthewgenchev/Desktop/USC/qbio_490_matthewgenchev/analysis_data/brca_rna_genes_data.csv")
rna_clinical <- read.csv("/Users/matthewgenchev/Desktop/USC/qbio_490_matthewgenchev/analysis_data/brca_rna_clincial_data.csv")

```

###### (1) Setting up and preprocessing our data

Firstly, we must convert the age_category column to a factor. Recall from the intro to R notebook that a factor is a categorical data type that stores values as well as different levels (a list of unique values). For example, within a column called 'grade' we might have the levels 'Freshman', 'Sophomore', 'Junior', 'Senior'.

Use factor() to convert rna_clinical$age_category into a factor.

```{r}
?factor

# write code here
rna_clinical$age_category <- factor(rna_clinical$age_category)

```

After factor conversion, check the levels by printing out the following:

```{r}
head(rna_clinical$age_category)
```

Note that in addition to the first few values, we now also see "Levels: old young". This indicates that rna_clinical$age_category was successfully converted into a factor.

In our analysis we can also control for variables that may skew our comparison. Let's control for stage at diagnosis (using ajcc_pathologic_stage), and one other variable of your choosing.

List your chosen variable here: prior_malignancy

Just like we did to age_category, convert these columns into the factor data type. Then call head and record the levels.

```{r}
rna_clinical$ajcc_pathologic_stage <- factor(rna_clinical$ajcc_pathologic_stage)
rna_clinical$prior_malignancy <- factor(rna_clinical$prior_malignancy)

head(rna_clinical$ajcc_pathologic_stage)
head(rna_clinical$prior_malignancy)
```

For DESeq2 to run, none of our variables can contain NAs. Just to see if there are NAs in age_category, ajcc_pathologic_stage, and your chosen variable by calling sum() on is.na() on the column.

```{r}
# write code here
sum(is.na(rna_clinical$ajcc_pathologic_stage))
sum(is.na(rna_clinical$prior_malignancy))
sum(is.na(rna_clinical$age_category))
```

If you get a non-zero number (aka there is at least one NA value in the column), create a boolean mask and get rid of any patients with an NA value. Make sure you delete these patients from both the rna_clinical data frame and the rna_counts data frame.

My rna_counts and rn_clinical are not the same length for patients so I am finding what patients are not in my clinical and getting rid of them in my rna_counts df
```{r}

col_name_mask <- colnames(rna_counts) %in% rownames(rna_clinical)

FALSE %in% col_name_mask

rna_counts <- rna_counts[ , col_name_mask]
```


```{r}
na_mask <- ifelse(is.na(rna_clinical$ajcc_pathologic_stage)
                  | is.na(rna_clinical$prior_malignancy)
                  | is.na(rna_clinical$prior_malignancy)
                  , F, T)
# you can use multiple NA masks if you have multiple columns with NA values (or you can do it all in one!)
  
rna_clinical <- rna_clinical[na_mask, ] # think about where the mask goes in the df[rows, columns] notation

rna_counts <- rna_counts[ , na_mask]  # does the mask go in the same or opposite location?
```

Now, let's move on to processing the rna_counts df. Here, we should remove all genes where the total number of counts (across all patients) is less than 10.

```{r}
```

```{r}
# use rowSums() to create a list with the total number of counts of each gene
row_sums <- rowSums(rna_counts[, c(-1, -2)])

# create a boolean mask where genes with < 10 total counts are FALSE, and genes with >= 10 total counts are TRUE
low_counts_mask <- ifelse(row_sums < 10, F, T)

# rewrite the rna_counts df, subsetting for only genes with >= 10 total counts
rna_counts <- rna_counts[low_counts_mask, ]

#update rna_genes with the low_counts_mas
rna_genes <- rna_genes [low_counts_mask, ]

#Removing text in the df
rna_counts <- rna_counts[ , -1:-2]
rna_clinical <- rna_clinical[-1:-2, ]
```

Now we're ready for the analysis!

###### (2) Running DESeq2

To break down what analysis actually takes place within the DESeq2 pipeline:

1. The DESeqDataSetFromMatrix() function processes your transcript count data (rna_counts), the phenotypic data (rna_clinical), and the categories you want to compare on (the design argument) to create a DESeqDataSet object (dds). In a design formula, the last variable is the condition you want to perform DE on. Optionally, you can also control for additional covariates by including them prior to your condition (see R block below).  

2. The DESeq() function runs the comparisons and statistical analysis. This will likely take the bulk of the run time. Don't worry if this takes quite a while to run, just let it take its time.

3. The resultsNames() function will tell you what comparisons have been run. It’s helpful for when you have more than two levels in your experiment (in our case, there are just two, "young" and "old").

4. The results() function will make the comparisons readable. The contrasts argument calculates the fold change in "young" vs. "old", with the second level in the list being the denominator in the fold change (ie the baseline for the expression calculations).

```{r}
#For some reason, my rna_counts data had strings in the 1st two columns so I just deleted the associated columns and rows from both dataframes
#rna_counts <- rna_counts [ , c(-1, -2)]
#rna_clinical <- rna_clinical [c(-1, -2), ]

#?DESeqDataSetFromMatrix
dds <- DESeqDataSetFromMatrix(countData = rna_counts,
                             colData = rna_clinical,
                             design = ~ajcc_pathologic_stage + prior_malignancy + age_category)

#?DESeq
dds_obj <- DESeq(dds) # note: this will likely take a long time (ie 45 minutes to 2 hours)

#?resultsNames

resultsNames(dds_obj)  # see what comparisons got run

# get the young vs. old comparison
#?results
results <- results(dds_obj, format = "DataFrame", contrast = c("age_category", "young", "old")) # this is case sensitive so be careful to match it with your age_category factors closely!
```

###### (3) Analyzing and saving the data

First, explore the results df using head(), View(), and dim().

```{r}
#head(results)
#dim(results)
#View(results)
```

We're not interested in all of these columns, so use data.frame() to write results with the following columns in this order: gene name (from rna_genes), gene id (from results rownames), log2foldchange, pvalue, padj, -log10(padj).

There will likely be fewer genes in results than in your rna_se based dataframes. That's okay, just use a mask to include only the genes that are in results. 

Rename the columns using colnames(). Note: to get -log10, simply write "-log10(fill in here)".

```{r}
# write code here, replace column#'s
modified_results <- data.frame(rna_genes$gene_name, 
                               results@rownames, 
                               results@listData$log2FoldChange, 
                               results@listData$pvalue,
                               results@listData$padj,
                               -log10(results@listData$padj))

# HINT: to get the gene names, you will need to create a boolean mask that is T when rna_genes$gene_id %in% results@rownames, and F if otherwise, then use this mask to subset rna_genes$gene_names for only the names that are in the results df
names_mask <- ifelse(rna_genes$gene_id %in% results@rownames, T, F)
modified_results <- modified_results[names_mask, ]

colnames(modified_results) <- c("gene_name", "ensembl", "log2_fold_change", "p_value", "p_adjusted", "-log10_p_adjusted") ## FIX
```

It's oftentimes most interesting to look at genes that are highly over/under expressed at a significance level p < 0.05. 

First, select rows (genes) that have a padj value < 0.05.

```{r}
sig_results_mask <- ifelse(modified_results$p_adjusted < 0.05, T, F)
modified_results <- modified_results[sig_results_mask, ]
```

Now, sort the data frame by log2foldchange, descending. To do this, use the order() function. The order() function returns a vector of the sorted indexes of your chosen data, from smallest to largest. For example:

```{r EXAMPLE SHOWING HOW ORDER() WORKS}
my_example_df = data.frame(x = c('index 1', 'index 2', 'index 3', 'index 4', 'index 5'),
                   y = c(2,4,3,5,1))

order_indexes = order(my_example_df$y) # order by column "y"

# we would expect order_indexes to be c(5, 1, 3, 2, 4) since the value 2 has the index 5, the value 2 has the index 1, the value 3 has the index 3, etc.

order_indexes

my_example_df = my_example_df[order_indexes, ] # this rewrites the df based on the sorted rows
my_example_df # check your df to see if it is sorted by the y values
```

Use the logic above as well as sig_results to create a dataframe called up_reg_results. This df should be sorted by log2foldchange descending (take a look at the decreasing argument). Note that this can also be done in one line by bypassing the use of order_indexes as an intermediate variable.

```{r}
order_indexes = order(-modified_results$log2_fold_change)
up_reg_results <- modified_results[order_indexes, ]
```

Rewrite up_reg_results to have only the genes where the log2foldchange is > 1 (ie genes that are 2x or greater expressed in one population over the other).

```{r}
# write code here
log2foldchange_mask <- ifelse(up_reg_results$log2_fold_change > 1, T, F)
up_reg_results <- up_reg_results[log2foldchange_mask, ]
up_reg_results <- up_reg_results[1:41, ] #For some reason, my dataframe was not shorter after the mask application

```

Repeat the same steps to create a sorted data frame called down_reg_results (this time select for all genes where log2foldchange is < -1 and sort it ascending by log2foldchange).

```{r}
# write code here
new_order_indexes = order(modified_results$log2_fold_change)
down_reg_results <- modified_results[new_order_indexes, ]
neg_log2foldchange_mask <- ifelse(down_reg_results$log2_fold_change < -1, T, F)
down_reg_results <- down_reg_results[neg_log2foldchange_mask, ]
down_reg_results <- down_reg_results[1:286, ] #Again, my code did not erase na values for some reason
```

Call head() on both up_reg_results and down_reg_results.

```{r}
# write code here
head(c(up_reg_results, down_reg_results))
```

Notice that the row names are the old indexes from before we sorted. Create new row names that are more descriptive (ie. gene_id). 

```{r}
# write code here
row.names(up_reg_results) <- up_reg_results$ensembl
row.names(down_reg_results) <- down_reg_results$ensembl
```

Call head() again, and choose one up regulated (overexpressed) gene and one down regulated (underexpressed) gene and fill out the following information:

Gene ensembl id: ENSG00000163631.17
Log2FoldChange: 2.546894
p-adj value: 1.117046e-32
More highly expressed in young or old patients: More highly expressed in young patients
Google the gene. What is its full name, and what appears to be its function? Genecards generally has a pretty good summary: The gene is Albumin (ALB), and is one of the most abundant proteins in the blood. It regulates blood plasma colloid osmotic pressure and acts as a carrier protein for a lot of endogenous molecules

Gene ensembl id: ENSG00000202354.1
Log2FoldChange: -2.817995
p-adj value: 3.420006e-02
More highly expressed in young or old patients: More expressed in old patients
Google the gene. What is its full name, and what appears to be its function? Genecards generally has a pretty good summary: RNY3; it is a small cytoplasmic ribonucleoprotein RY3,component of Ro ribonucleoprotein

Write the results, up_reg_results, and down_reg_results df's to analysis_data using write.csv().

```{r}
# write code here
write.csv(up_reg_results, "/Users/matthewgenchev/Desktop/USC/qbio_490_matthewgenchev/analysis_data/up_regulated_results.csv", row.names = TRUE)

write.csv(down_reg_results, "/Users/matthewgenchev/Desktop/USC/qbio_490_matthewgenchev/analysis_data/down_regulated_results.csv", row.names = TRUE)
```

###### (4) Creating volcano plots

The most common way to visualize differential expression results is with a volcano plot (named because it looks like a volcano erupting). A volcano plot shows the relative expression of each gene along with the significance for each gene's result.

The x-axis is log2FoldChange (ie the normalized fold change results), and the y axis is the -log10(p adjusted value).

First, think conceptually about which points in the volcano plot are the most interesting. Fill out the relative expression (up-regulated/down-regulated in young patients) and significance (yes/no) for the following locations on the volcano plot:

top-right genes: up-regulated in young, significant
bottom-right genes: up-regulated in young, not significant
top-left genes: down-regulated in young, significant
bottom-left genes: down-regulate in young, not significant
top-middle genes: Not up/down regulated in young, significant
bottom-middle genes: Not up/down regulated in young, not significant

There are many general plotting functions you can use to create a volcano plot (ex: plot(), gglot(), etc.), but we are going to use a specialized function made particularly for volcano plots as it allows for more customization. Install and load "EnhancedVolcano".

```{r}
# install (using BiocManager) and load "EnhancedVolcano"
BiocManager::install("EnhancedVolcano")
library(EnhancedVolcano)
```

Read the package documentation section 3.1, and create a volcano plot of your own.
https://www.bioconductor.org/packages/release/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html#plot-the-most-basic-volcano-plot
```{r}
modified_na_mask <- ifelse(is.na(modified_results$gene_name), F, T)
modified_results <- modified_results[modified_na_mask, ]
row.names(modified_results) <- modified_results$ensembl
```

Remember that you may need to include par(mar=c(1,1,1,1)) before your plotting function to change the margin sizes.

```{r}
# create plot here
par(mar=c(1,1,1,1))
EnhancedVolcano(modified_results,
                lab = rownames(modified_results),
                x = 'log2_fold_change',
                y = 'p_adjusted')
```

Skim through section 4 of the package documentation, and create an advanced volcano plot as you see fit. Customize at least one aspect of your plot (ie color, shape, labels, etc.). Save this plot to your week8 folder as a .jpg file. For more info on saving plots, see section 3 of R_Tutorial.Rmd

```{r}
par(mar=c(1,1,1,1))
EnhancedVolcano(modified_results,
                title = "Gene Expression in Young vs Old Patients",
                lab = rownames(modified_results),
                x = 'log2_fold_change',
                y = 'p_adjusted',
                labSize = 2,
                boxedLabels = TRUE,
                drawConnectors = TRUE,
                max.overlaps = 20,
                shape = 13,
                col = c('darkgreen', 'navy', 'peru', 'plum1' ),
                colAlpha = 0.5)

```

